### 2015-10-09 manuscript.Rnw
various edits

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore research questions in evolution, biodiversity, and conservation, the
resources OTL provides could be used for education and outreach (e.g.,
illustrating course material, develop outreach activities to explore
relationships among species).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
Application Programming Interface (API) to the different services to work
programmatically with the underlying data. Here, we present \texttt{rotl} an R
package to interact with OTL's API using R syntax to search for, and import
phylogenetic trees and taxonomic information into R. Having access to this data
directly in R allows researchers to take advantage of the multiple packages that
are available to visualize, analyze and manipulate phylogenetic and comparative
data (ape, phytools, geiger, ggtree, phylobase, RNeXML, etc... cite task view
\url{https://cran.r-project.org/web/views/Phylogenetics.html}).

\section{Services provided by OTL}

The OTL project provides four resources that serve data to users through the
API:

\begin{enumerate}
\item The \emph{taxonomy} used as the backbone from the tree, the Open Tree
  Taxonomy;
\item The \emph{studies} and their associated trees some of which are chosen by
  curators to assemble the synthetic tree;
\item A \emph{taxonomic name resolution service} (TNRS) used to match taxon
  names to the Open Tree Taxonomy identifiers;
\item The \emph{synthetic tree} itself (the Open Tree).
\end{enumerate}

Additionally, the project provides access to other data, the Graph of Life, that
stores information about the underlying tree alignment graphs \cite{Smith2013}
used to synthesize the various sources of information from which the Open Tree
is assembled from. Data from this API endpoint is not intended for normal users
and its use will not be demonstrated in this paper but \texttt{rotl} has
functions to retrieve data it provides.

@@

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by the API. However, resources and methods are being developed to
add branch lengths information to these topological trees
\citep[e.g.,][]{Ksepka2015} or use topological trees to identify
phylogenetically equivalent species to increase overlap between chronograms and
species trait data \citep{Pennell2015}. Without branch lengths, these trees
remain useful to illustrate relationships among species, or to map a trait on a
phylogeny for instance.

@@

Phylogenetic information retrieved from OTL is converted into
\texttt{ape::phylo} objects by \texttt{rotl} using the NEXUS Class Library
\citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
\url{https://cran.r-project.org/package=rncl}. Using NCL provides robust and
efficient parsing of large trees that may contain singletons nodes labeled with
taxonomic information.

@@

The function \texttt{tnrs\_match\_names} returns a data frame that lists the
Open Tree identifiers as well as other information to help users to ensure that
the taxa matched are the correct ones. Here, there is no ambiguity in the taxa
matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
and animals that are regulated by different nomenclatural codes (ICNP, ICN and
ICZN respectively), OTL and \texttt{rotl} provide tools to deal with potential
hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
matches to a taxonomic group (see the function \texttt{tnrs_contexts} for a list
of possible options).  When this strategy cannot be used in case the tree
encompasses multiple domains (as in the present example), the function
\texttt{inspect} lists alternative matches for a taxon name, and \texttt{update}
replaces it in the results. An example of this approach is provided in the
vignette ``How to use rotl?'' that accompanies the package.

@@

When more than one tree is available for a given study, the function
\texttt{list\_trees} returns a list containing the tree identifiers for each
study easily. Alternatively, the function \texttt{get\_study} returns all the
trees (by default as \texttt{ape::phylo} objects) associated with a particular
study. Additional details about the study (i.e., its metadata) can be obtained
using the function \texttt{get\_study\_meta}.

@@

In recent years, R has become part of the toolbox of many researchers in
evolutionary biology. As many methods for comparative analyses are implemented,
and because it is a relatively easy to use programming language, R greatly
facilitates the analysis of large datasets, and allows researchers to combine
methods in novel ways. Additionally, as more data is made available online and
accessible using APIs, many R packages have been developed to take advantage of
these datasets, especially spearheaded by the organization rOpenSci
\url{https://ropensci.org}. \texttt{rotl} contributes to these initiatives and
data from OTL can be combined with other sources easily. For instance in
Figure~\ref{cat_map}, we show how we can obtain a map of the occurrences for
some of the cat species (genus \textit{Lynx}) included in the phylogeny above
using records for these species found in GBIF (code is available in
supplementary material).

<<cat_map, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=

@@

lynx_loc <- do.call("rbind", lynx_loc)
lynx_loc[["decimalLatitude"]] <- as.numeric(lynx_loc[["decimalLatitude"]])
lynx_loc[["decimalLongitude"]] <- as.numeric(lynx_loc[["decimalLongitude"]])
lynx_loc <- lynx_loc[complete.cases(lynx_loc), ]

>>>>

The Open Tree of Life (OTL) project aims at assembling and synthesizing our
current understanding of phylogenetic relationships across all organisms on
Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
access to this information. OTL combines taxonomic information that serves as
the backbone for the phylogenetic relationships, and published phylogenies to
elucidate relationships among taxa. This combination of information is used to
form the synthetic tree. Studies can be contributed to the synthetic tree
through the curator interface \url{https://tree.opentreeoflife.org/curator},
allowing the synthetic tree to be continuously updated as relationships are
elucidated or reevaluated. The current draft of the Open Tree of Life contains
2.3 million tips. Beyond obvious applications across the life sciences to
explore questions in evolution, biodiversity, and conservation, the resources
OTL provides could be used for education and outreach (e.g., illustrating course
material, develop outreach activities to explore relationships among species).

@@

The Open Tree of Life project provides a web interface to explore, query and
download parts of the synthetic Open Tree
(\url{http://tree.opentreeoflife.org}). Additionally, the project provides
Application Programming Interfaces (APIs) to the different services to work
programmatically with the underlying data. Here, we present \texttt{rotl} an R
package to interact with OTL's APIs using R syntax to search for, and import
phylogenetic trees and taxonomic information into R. Having access to this data
directly in R allows researchers to take advantage of the multiple packages that
are available to visualize, analyze and manipulate phylogenetic and comparative
data (e.g., ape, phytools, geiger, ggtree, phylobase, RNeXML, see
\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
comprehensive list). Furthermore, \texttt{rotl} joins a growing list of R
packages that aim at accessing data from the web.

\section{API services provided by OTL}

The OTL project provides four resources that serve data to users through the
APIs:

\begin{enumerate}
\item The \emph{taxonomy} used as the backbone from the tree, the Open Tree
  Taxonomy;
\item The \emph{studies} and their associated trees some of which are chosen by
  curators to assemble the synthetic tree;
\item A \emph{taxonomic name resolution service} (TNRS) used to match taxon
  names to the Open Tree Taxonomy identifiers;
\item The \emph{synthetic tree} itself, the Open Tree.
\end{enumerate}

Additionally, the project provides access to other data, the Graph of Life, that
stores information about the underlying tree alignment graphs \citep{Smith2013}
used to synthesize the various sources of information from which the Open Tree
is assembled from. Data from this API is not intended for normal users, and its
use will not be demonstrated in this paper but \texttt{rotl} has functions to
retrieve data it provides.

@@

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by OTL. However, resources and methods are being developed to add
branch lengths information to these topological trees
\citep[e.g.,][]{Ksepka2015} or use topological trees to identify
phylogenetically equivalent species to increase overlap between chronograms and
species trait data \citep{Pennell2015}. Without branch lengths, these trees
remain useful to illustrate relationships among species, or to map a trait on a
phylogeny for instance.

@@

Phylogenetic information retrieved from OTL is converted into
\texttt{ape::phylo} objects by \texttt{rotl} using the NEXUS Class Library
\citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
\url{https://cran.r-project.org/package=rncl}. Using NCL provides robust and
efficient parsing of large trees that may contain singleton nodes labeled with
taxonomic information.

@@

The function \texttt{tnrs\_match\_names} returns a data frame that lists the
Open Tree identifiers as well as other information to help users to ensure that
the taxa matched are the correct ones. Here, there is no ambiguity in the taxa
matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
and animals that are regulated by different nomenclatural codes (ICNP, ICN and
ICZN respectively), OTL and \texttt{rotl} provide tools to deal with potential
hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
matches to a taxonomic group (see the function \texttt{tnrs\_contexts} for a list
of possible options).  When this strategy cannot be used in case the tree
encompasses multiple domains (as in the present example), the function
\texttt{inspect} lists alternative matches for a taxon name, and \texttt{update}
replaces it in the results. An example of this approach is provided in the
vignette ``How to use \texttt{rotl}?'' that accompanies the package.

@@

When more than one tree is available for a given study, the function
\texttt{list\_trees} returns a list containing the tree identifiers for each
study. Alternatively, the function \texttt{get\_study} returns all the trees (by
default as \texttt{ape::phylo} objects) associated with a particular
study. Additional details about the study (i.e., its metadata) can be obtained
using the function \texttt{get\_study\_meta}.

@@

In recent years, R has become part of the toolbox of many researchers in
evolutionary biology. As many methods for comparative analyses are implemented,
and because it is a relatively easy to use programming language, R greatly
facilitates the analysis of large datasets, and allows researchers to combine
methods in novel ways. Additionally, as more data is made available online and
accessible using APIs, many R packages have been developed to take advantage of
these datasets, especially spearheaded by the organization rOpenSci
\url{https://ropensci.org}. \texttt{rotl} contributes to these initiatives and
data from OTL can be combined with other sources easily. For instance in
Figure~\ref{fig:cat_map}, we show how we can obtain a map of the occurrences for
some of the cat species (genus \textit{Lynx}) included in the phylogeny above
using records for these species found in GBIF (code is available in
supplementary material).

<<cat_map, fig.height=4.2, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=

@@

lynx_loc <- do.call("rbind", lynx_loc)
lynx_loc[["decimalLatitude"]] <- as.numeric(lynx_loc[["decimalLatitude"]])
lynx_loc[["decimalLongitude"]] <- as.numeric(lynx_loc[["decimalLongitude"]])
lynx_loc[lynx_loc[["decimalLatitude"]] == 0 &
         lynx_loc[["decimalLongitude"]] == 0,
         c("decimalLatitude","decimalLongitude")] <- c(NA, NA)
lynx_loc <- lynx_loc[complete.cases(lynx_loc), ]
